|

### 2021-07-30 prova_knit_finale.Rmd
intro added
SVAPFopt è errata

\subsection{Implementation}
\
Consider the linear Gaussian example of section \ref{implementation}, but this time with unknown variances $\tau^{2}$ and $\sigma^{2}$. Thus, let $\psi=(\sigma^{2},\tau^{2})$ be the unknown parameter vector and assign a gamma prior for its components,
\begin{align*}
\sigma^{2}  & \sim G(\alpha_{v},\beta_{v}) \\
\tau^{2}  & \sim G(\alpha_{w},\beta_{w})
\end{align*}
Alternatively, assign them a uniform prior if we have no knowledge of the hyperparameters. Note that in this model we sample from a Gamma distribution (instead of the Normal one, as in the previous description of the LWF) since the parameters of interest are variances, that can take only positive values. The algorithm follows the following steps.
\begin{algorithm} LWF for Random Walk plus Noise Model
\begin{itemize}
\item Initialize $(x_{0}^{(1)},...,x_{0}^{(N)})$ from $N(m_{0},C_{0})$, $({\sigma^{2}}^{(1)},...,{\sigma^{2}}^{(N)})$ from $G(\alpha_{v},\beta_{v})$ and $({\tau^{2}}^{(1)},...,{\tau^{2}}^{(N)})$ from $G(\alpha_{w},\beta_{w})$. Set $w_{0}^{(i)}=N^{-1} \ \forall \ i=1,...,N$. Therefore $\psi^{(i)}=({\sigma^{2}}^{(i)},{\tau^{2}}^{(i)})$, and
$\hat{\pi}_{0}=p(x_{0}|y_{0})=\sum_{i=1}^{N}w_{0}^{(i)}\delta_{(x_{0}^{(i)},\psi^{(i)})}$
\item For $t=1,...,n$:.
\begin{enumerate}
\item Compute $\overline{\psi}=E_{\hat{\pi}_{t-1}}(\psi)$ and $\Sigma=V_{\hat{\pi}_{t-1}}(\psi)$. For $i=1,...,N$, set
\begin{align*}
m^{(i)} & = a\psi^{(i)}+(1-a)\hat{\psi} \\
\hat{x}_{t}^{(i)} & = E(x_{t}|x_{t-1}=x_{t-1}^{(i)},\psi=\psi^{(k)})
\end{align*}
\item For $k=1,...,N$:
\begin{itemize}
\item Draw $I_{k}$, with $P(I_{k}=i) \propto w_{t-1}^{(i)}f_{N}(y_{t}|g(x_{t-1}^{(i)}),\psi=m^{(i)}) $
\item Draw ${\sigma^{2}}^{(k)}$ from $G(\alpha_{v}^{(I_k)},\beta_{v}^{(I_k)})$ 
\item Draw ${\tau^{2}}^{(k)}$ from $G(\alpha_{w}^{(I_k)},\beta_{w}^{(I_k)})$
\item Draw $x_{t}^{(k)}$ from $N(x_{t-1}^{(I_k)},\psi=\psi^{(k)})$
\item Set  $\tilde{w}_{t}^{(k)} = \frac{f_{N}(y_{t}|x_{t}^{(k)},\psi=\psi^{(k)})}{f_{N}(y_{t}|g(x_{t-1}^{(I_{k})},\psi=m^{(I_k)}))}$
\end{itemize}
\item Compute $ESS=\Bigg(\sum_{i=1}^{N}(w_{t}^{(i)})^{2}\Bigg)^{-1}$
\item if $ESS<N/2$ then
\begin{enumerate}
\item Draw a sample of size N, $(x_{t}^{(1)},...,x_{t}^{(N)})$, from the discrete distribution $P(x_{t}=x_{t}^{(i)})=w_{t}^{(i)},\ \ i=1,...,N$
\item Reset the weights: $w_{t}^{(i)}=N^{-1}$, $i=1,...,N$.
\end{enumerate}
\item Set $\hat{\pi}_{t}=p(x_{t}|y_{1:t})=\sum_{i=1}^{N}w_{t}^{(i)}\delta_{(x_{t}^{(i)},\psi^{(i)})}$
\end{enumerate}
\end{itemize}
\end{algorithm}
Our \code{LWfun} function goes through the illustrated steps.
\
\hrule
\hrule
\
\code{LWfun(data,N,m0,C0,alphav,betav,alphaw,betaw,delta,unif,r)}
\
\hrule
\textbf{Arguments}

@@

So far we have seen Sequential Monte Carlo methods applied to the linear Gaussian case, however, as already mentioned, for this class of state space models, Kalman Filter provides a closed form solution

@@

    xprev<-x
    x<-rnorm(N,alpha+beta*x+(tau^2)/4*((data[t]^2)*exp(-(alpha+beta*x))-2),tau)
    w1<-w*dnorm(data[t],0,exp(x/2))*dnorm(x,alpha+beta*xprev,tau)/dnorm(x,mean=alpha+beta*xprev+(tau^2)/4*((data[t]^2)*exp(-(alpha+beta*xprev))-2),tau)

@@

```{r}
###################
## Per roberto:::::: cambia molto se usiamo l'uno o l'altro!!!!!! perchè??
APFoptfun<-function(data,N,m0,C0,tau,sigma,r){
  if(missing(r)){r=2}else{}
  xs<-NULL
  ws<-NULL
  ess<-NULL
  x  = rnorm(N,m0,sqrt(C0))
  importancesd<-sqrt(tau - tau^2 /(tau + sigma))
  predsd <- sqrt(sigma+tau)
  w  = rep(1/N,N)

    weight = w*dnorm(data[t],x,predsd)
    k   = sample(1:N,size=N,replace=TRUE,prob=weight)
    means<-x[k]+(tau/(tau+sigma))*(data[t]-x[k])
    x1   = rnorm(N,means,importancesd)
    lw  = dnorm(data[t],x1,predsd,log=TRUE)-dnorm(data[t],x[k],predsd,log=TRUE)+log(w)-log(weight)
    w   = exp(lw)
    w   = w/sum(w)
    ESS  = 1/sum(w^2)

  }
  return(list(xs=xs,ws=ws,ess=ess))
}
```
\
\hrule
\hrule
\
\code{SISfun(data,N,m0,C0,alpha,beta,tau,r)}
\
\hrule
\textbf{Arguments}

>>>>

\subsection{Implementation}
\
Consider the linear Gaussian example of section \ref{implementation}, but this time with unknown variances $\tau^{2}$ and $\sigma^{2}$. Thus, let $\psi=(\sigma^{2},\tau^{2})$ be the unknown parameter vector and assign a gamma prior for its components,
\begin{align*}
\sigma^{2}  & \sim G(\alpha_{v},\beta_{v}) \\
\tau^{2}  & \sim G(\alpha_{w},\beta_{w})
\end{align*}
Alternatively, assign them a uniform prior if we have no knowledge of the hyperparameters. Note that in this model we sample from a Gamma distribution (instead of the Normal one, as in the previous description of the LWF) since the parameters of interest are variances, that can take only positive values. Next, we descrive the algorithm.
\begin{algorithm} LWF for Random Walk plus Noise Model
\begin{itemize}
\item Initialize $(x_{0}^{(1)},...,x_{0}^{(N)})$ from $N(m_{0},C_{0})$, $({\sigma^{2}}^{(1)},...,{\sigma^{2}}^{(N)})$ from $G(\alpha_{v},\beta_{v})$ and $({\tau^{2}}^{(1)},...,{\tau^{2}}^{(N)})$ from $G(\alpha_{w},\beta_{w})$. Set $w_{0}^{(i)}=N^{-1} \ \forall \ i=1,...,N$. Therefore $\psi^{(i)}=({\sigma^{2}}^{(i)},{\tau^{2}}^{(i)})$, and
$\hat{\pi}_{0}=p(x_{0}|y_{0})=\sum_{i=1}^{N}w_{0}^{(i)}\delta_{(x_{0}^{(i)},\psi^{(i)})}$
\item For $t=1,...,n$:.
\begin{enumerate}
\item Compute $\overline{\psi}=E_{\hat{\pi}_{t-1}}(\psi)$ and $\Sigma=V_{\hat{\pi}_{t-1}}(\psi)$. For $i=1,...,N$, set
\begin{align*}
m^{(i)} & = a\psi^{(i)}+(1-a)\hat{\psi} \\
\hat{x}_{t}^{(i)} & = E(x_{t}|x_{t-1}=x_{t-1}^{(i)},\psi=\psi^{(k)})
\end{align*}
\item For $k=1,...,N$:
\begin{itemize}
\item Draw $I_{k}$, with $P(I_{k}=i) \propto w_{t-1}^{(i)}f_{N}(y_{t}|g(x_{t-1}^{(i)}),\psi=m^{(i)}) $
\item Draw ${\sigma^{2}}^{(k)}$ from $G(\alpha_{v}^{(I_k)},\beta_{v}^{(I_k)})$ 
\item Draw ${\tau^{2}}^{(k)}$ from $G(\alpha_{w}^{(I_k)},\beta_{w}^{(I_k)})$
\item Draw $x_{t}^{(k)}$ from $N(x_{t-1}^{(I_k)},\psi=\psi^{(k)})$
\item Set  $\tilde{w}_{t}^{(k)} = \frac{f_{N}(y_{t}|x_{t}^{(k)},\psi=\psi^{(k)})}{f_{N}(y_{t}|g(x_{t-1}^{(I_{k})},\psi=m^{(I_k)}))}$
\end{itemize}
\item Compute $ESS=\Bigg(\sum_{i=1}^{N}(w_{t}^{(i)})^{2}\Bigg)^{-1}$
\item if $ESS<N/2$ then
\begin{enumerate}
\item Draw a sample of size N, $(x_{t}^{(1)},...,x_{t}^{(N)})$, from the discrete distribution $P(x_{t}=x_{t}^{(i)})=w_{t}^{(i)},\ \ i=1,...,N$
\item Reset the weights: $w_{t}^{(i)}=N^{-1}$, $i=1,...,N$.
\end{enumerate}
\item Set $\hat{\pi}_{t}=p(x_{t}|y_{1:t})=\sum_{i=1}^{N}w_{t}^{(i)}\delta_{(x_{t}^{(i)},\psi^{(i)})}$
\end{enumerate}
\end{itemize}
\end{algorithm}
Our \code{LWfun} function goes through the illustrated steps.
\
\hrule
\hrule
\
\code{LWfun(data,N,m0,C0,alphav,betav,alphaw,betaw,delta,unif,r)}
\
\hrule
\textbf{Arguments}

@@

So far we have seen Sequential Monte Carlo methods applied to the linear Gaussian case, however, as already mentioned, for this class of state space models, Kalman Filter provides a closed form solution. Here, we want to provide an illustration of Particle Filters for a state space model for which no analytical solutions are available. More precisely, we deal with a stochastic volatility model. For the sake of simplicity we use a basic specification of stochastic volatility model

\begin{align}
y_{t}|x_{t} & \sim N(0,e^{x_{t}}) \\
x_{t}|x_{t-1} & \sim N(\alpha+\beta x_{t-1},\tau^2)
\end{align}

We start describing the main four particle filtering strategies discussed in this work and then we are going to compare their performances. The R code of these and other algorithms used in the Chapter application can be found in Appendix 1.

\begin{algorithm} BPF for Stochastic Volatility Model
\begin{itemize}
\item Let $\{(x_{0},w_{0})^{(i)}\}_{i=1}^{N}$ summarize $p(x_{0}|y_{0})$ such that, for example, $E(g(x_{0})|y_{0}) \approx \sum_{i=1}^{N}w_{0}^{(i)}g(x_{0}^{(i)})$. In particular, initialize $(x_{0}^{(1)},...,x_{0}^{(N)})$ form $N(m_{0},C_{0})$ and set $w_{0}^{(i)}=N^{-1} \ \forall \ i=1,...,N$.
\item For $t=1,...,n$:
\begin{enumerate}
\item Draw $x_{t}^{(i)} \sim N(\alpha+\beta x_{t-1},\tau^2) \ \ i=1,...,N$ such that $\{(x_{t},w_{t-1})^{(i)}\}_{i=1}^{N}$ summarizes $p(x_{t}|y_{1:t-1})$
\item Set $w_{t}^{(i)} = w_{t-1}^{(i)}f_{N}(y_{t};0,e^{x_{t}}) \ \ i=1,...,N$ such that $\{(x_{t},w_{t})^{(i)}\}_{i=1}^{N}$ summarizes $p(x_{t}|y_{1:t})$
\item Normalize the weights: $w_{t}^{(i)}=\frac{\tilde{w}_{t}^{(i)}}{\sum_{j=1}^{N}(\tilde{w}_{t}^{(j)})}$
\item Compute $ESS=\Bigg(\sum_{i=1}^{N}(w_{t}^{(i)})^{2}\Bigg)^{-1}$
\item if $ESS<N/2$ then
\begin{enumerate}
\item Draw a sample of size N, $(x_{t}^{(1)},...,x_{t}^{(N)})$, from the discrete distribution $P(x_{t}=x_{t}^{(i)})=w_{t}^{(i)},\ \ i=1,...,N$
\item Reset the weights: $w_{t}^{(i)}=N^{-1}$, $i=1,...,N$.
\end{enumerate}
\item Set $p(x_{t}|y_{1:t})=\sum_{i=1}^{N}w_{t}^{(i)}\delta_{x_{t}^{(i)}}$
\end{enumerate}
\end{itemize}
\end{algorithm}

\begin{algorithm} GPF for Stochastic Volatility Model
\begin{itemize}
\item Let $\{(x_{0},w_{0})^{(i)}\}_{i=1}^{N}$ summarize $p(x_{0}|y_{0})$ such that, for example, $E(g(x_{0})|y_{0}) \approx \sum_{i=1}^{N}w_{0}^{(i)}g(x_{0}^{(i)})$. In particular, initialize $(x_{0}^{(1)},...,x_{0}^{(N)})$ from $N(m_{0},C_{0})$ and set $w_{0}^{(i)}=N^{-1} \ \forall \ i=1,...,N$.
\item For $t=1,...,n$:
\begin{enumerate}
\item Draw $x_{t}^{(i)} \sim N(\alpha+\beta x_{t-1}+\frac{\tau^2}{4}\big[y_{t}^2e^{-(\alpha+\beta x_{t-1})}-2\big],\tau^{2}) \ \ i=1,...,N$ such that $\{(x_{t},w_{t-1})^{(i)}\}_{i=1}^{N}$ summarizes $p(x_{t}|y_{1:t-1})$
\item Set $w_{t}^{(i)} = w_{t-1}^{(i)}\frac{f_{N}(y_{t};0,e^{x_{t}})f_{N}(x_{t};\alpha+\beta x_{t-1},\tau^{2})}{f_{N}(x_{t};\alpha+\beta x_{t-1}+\frac{\tau^2}{4}\big[y_{t}^2e^{-(\alpha+\beta x_{t-1})}-2\big],\tau^{2})} \ \ i=1,...,N$ such that $\{(x_{t},w_{t})^{(i)}\}_{i=1}^{N}$ summarizes $p(x_{t}|y_{1:t})$
\item Normalize the weights: $w_{t}^{(i)}=\frac{\tilde{w}_{t}^{(i)}}{\sum_{j=1}^{N}(\tilde{w}_{t}^{(j)})}$
\item Compute $ESS=\Bigg(\sum_{i=1}^{N}(w_{t}^{(i)})^{2}\Bigg)^{-1}$
\item if $ESS<N/2$ then
\begin{enumerate}
\item Draw a sample of size N, $(x_{t}^{(1)},...,x_{t}^{(N)})$, from the discrete distribution $P(x_{t}=x_{t}^{(i)})=w_{t}^{(i)},\ \ i=1,...,N$
\item Reset the weights: $w_{t}^{(i)}=N^{-1}$, $i=1,...,N$.
\end{enumerate}
\item Set $p(x_{t}|y_{1:t})=\sum_{i=1}^{N}w_{t}^{(i)}\delta_{x_{t}^{(i)}}$
\end{enumerate}
\end{itemize}
\end{algorithm}

\begin{algorithm}APF with optimal transition kernel for Stochastic Volatility Model
\begin{itemize}
\item Let $\{(x_{0},w_{0})^{(i)}\}_{i=1}^{N}$ summarizes $p(x_{0}|y_{0})$ such that, for example, $E(g(x_{0})|y_{0}) \approx \sum_{i=1}^{N}w_{0}^{(i)}g(x_{0}^{(i)})$. In particular, initialize $(x_{0}^{(1)},...,x_{0}^{(N)})$ from $N(m_{0},C_{0})$ and set $w_{0}^{(i)}=N^{-1} \ \forall \ i=1,...,N$.
\item Compute $\sigma_{opt}^{2}$ 
\item For $t=1,...,n$:
\begin{enumerate}
\item Compute $ESS=\Bigg(\sum_{i=1}^{N}(w_{t}^{(i)})^{2}\Bigg)^{-1}$
\item if $ESS<N/2$ then
\begin{enumerate}
\item Draw $I_{k}$ with $P(I_{k}=i) \propto w_{t-1}^{(i)}f_{N}(y_{t}|g(x_{t-1}^{(i)}),\sigma^{2}+\tau^{2}) $
\item Compute $\mu_{opt}=\mu(x_{t-1}^{(I_{k})})$
\item Draw $x_{t}^{(k)} \sim N(\mu_{opt},\sigma^{2}_{opt})$
\item Set  $\tilde{w}^{(i)}_{t} = N^{-1}$ for $i=1,...,N$
\end{enumerate}
\item otherwise
\begin{enumerate}
\item Compute $\mu_{opt}=\mu(x_{t-1})$
\item Draw $x_{t}^{(i)} \sim N(\mu_{opt},\sigma^{2}_{opt})$ for $i=1,...N$
\item Set $\tilde{w}^{(i)}_{t}=w^{(i)}_{t-1}f_{N}(y_{t};x_{t}^{(i)},\sigma^{2}+\tau^{2})$ for $i=1,...,N$
\end{enumerate}
\item Normalize the weights: $w_{t}^{(i)}=\frac{\tilde{w}_{t}^{(i)}}{\sum_{j=1}^{N}(\tilde{w}_{t}^{(j)})}$
\end{enumerate}
\end{itemize}
\end{algorithm}

@@

    xprev<-x
    x<-rnorm(N,alpha+beta*x+(tau^2)/4*((data[t]^2)*exp(-(alpha+beta*x))-2),tau)
    w1<-w*dnorm(data[t],0,exp(x/2))*dnorm(x,alpha+beta*xprev,tau)/
      dnorm(x,mean=alpha+beta*xprev+(tau^2)/4*((data[t]^2)*exp(-(alpha+beta*xprev))-2),tau)

@@

\
\hrule
\hrule
\
\code{SISfun(data,N,m0,C0,alpha,beta,tau,r)}
\
\hrule
\textbf{Arguments}
